Polymethylation scores paper

### 2021-10-26 Papers/Polymethylation_scores/06_discussion.Rmd

@@

Restate findings

How are results similar or different to other results

Tissue

Age

Ancestry

>>>>

---
output:
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
---

@@

In the longitudinal Fragile Families and Child Wellbeing birth cohort, we found an association between prenatal maternal smoking and DNA methylation in sequential saliva samples from ages 9 and 15. Polymethylation scores were associated with prenatal maternal smoking across ages and ancestry while global methylation and epigenetic clocks were not, demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation. Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking. This accuracy was more consistent across time than when using some individual *a priori* CpG sites.

Our findings fit within the previous literature on associations between prenatal maternal smoking and DNA methylation. We replicate top hits from previous epigenome wide association studies, including a large metaanalysis of DNA methylation in blood samples [@Joubert:2016ue]. Although there are no previous studies of prenatal maternal smoking and saliva methylation, our findings fit with findings on the consistency of DNA methylation signatures of own-smoking between saliva and blood samples [@Dawes2019; @Philibert2020].

We contribute to the literature on DNA methylation and prenatal maternal smoking by evaluating the consistency of this association in saliva and across ages and ancestries. Polymethylation scores built using coefficients from metaanalysis of blood DNA methylation still associated with prenatal maternal smoking in our independent population of saliva samples. Similarly, even though the source population for construction of the polymethylation scores was primarily of European ancestry, an association with prenatal maternal smoking was apparent in our primarily African ancestry population. Effect estimates were consistent across ancestries at age 9, although Hispanic and European ancestry age 15 samples had a lower effect estimate. This could reflect higher unreported smoking initiation in European and Hispanic ancestry US teens than African ancestry teens. Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016]. Own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification. This could also explain the observed attenuation of hypomethylation at AHRR:CpG05575921 by age 15 in our sample, as methylation at this probe is known to vary by own-smoking exposure as well[@Dawes2019; @Philibert2020].

Our findings also fit with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure. Accuracy of the polymethylation scores in our sample (Age 9: 0.67 and Age 15: 0.66) was worse than the performance of biomarkers using cord blood (AUCs ranging from 0.82 to 0.97) [@Reese:2017ts; @Ladd-Acosta:2016aa]. Differences in methylation patterns across tissues and over time could be responsible for this diminished accuracy. However, our accuracy was comparable to accuracy when using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu]. The diminished accuracy from cord blood may be an acceptable trade-off considering saliva is a more readily available tissue for epidemiological and clinical applications.

Our analysis answers pertinent questions for the development of a clinically applicable methylation biomarker of prenatal maternal smoking. We show that DNA methylation scores calculated from children's saliva can accurately classify prenatal maternal smoke exposure in non-European ancestry children. Other bomarkers and risk scores are known to differ across ancestry groups, potentially contributing to and exacerbating health disparities. As new biomarkers are developed, understanding their consistency across ancestry groups is critical. We also demonstrated that accuracy is not substantially affected when using coefficients from newborn cordblood (no cell type control, age 9: 0.62, age 15: 0.61) vs older children peripheral blood (no cell type control age 9: 0.63, age 15: 0.60). However, using coefficients from models that incorporated cell-type control at the time of calculation did result in better accuracy, suggesting cell-type control should be instituted both when creating weights and when applying scores to new population. Including samples from both ages did not improve accuracy, thus repeated samplings may be unnecessary for clinical applications. Additionally, accuracy was more consistent when using polymethylation scores than using the single top-hit from Joubert et al, AHHR:CpG05575921 [@Joubert:2016ue]. Salivary AHHR:CpG05575921 methylation was less accurate in our sample at predicting prenatal maternal smoke exposure than when it has been used to classify own smoking in blood or saliva [@Philibert2020]. While using a single CpG site as a biomarker may be clinically pragamatic, single sites may be less robust to time-since-exposure and new environmental exposures. Alternatively, incorporating information from across several CpGs in a polymethylation score may be more robust to these .

Our analysis is not without its limitations. We used maternal self report of prenatal maternal smoker as serum cotinine levels were not available. Due to social desirability bias, this could result in exposure misclassification. We would expect this to bias our results towards the null. In any analysis of a prenatal exposure and postnatal outcome, there is the possibility of selection bias into the cohort due to live birth bias. Selection bias is also possible due to loss-to-follow-up between birth and age 9. Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility of residual confounding in this observational cohort. However, the longitudinal design with repeated measures, exposure-outcome temporality and large sample size are strengths of our analysis.

In total, our analysis contributes to the clinical implementation of a DNA methylation-based biomarker of prenatal maternal smoking. We showed that DNA methylation in saliva, a clinically practical tissue, had strong associations with and reasonable accuracy for prenatal maternal smoke exposure. Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for health equity. The development and application of biomarker for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice. Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible. Given the prevalence and negative health consequences of prenatal maternal smoke exposure, such a biomarker could be used to identify exposed individuals who might especially benefit from certain interventions.

### 2021-10-26 Papers/Polymethylation_scores/08_methods.Rmd
work on intro and background outline

---
output: html_document
bibliography: "../Methylation_SmokingBib.bib"  
---

# Methods

## Cohort

The Fragile Families and Child Wellbeing Study (FFCW) is a nationally-representative cohort of nearly 5,000 children born in 20 U.S. cities between 1998 and 2000 [@Reichman:2001wv]. FFCW sampled births using a three-stage stratified random sample design, described in detail elsewhere [@Reichman:2001wv]. FFCW excluded the following births from the study: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, those where the mothers or babies were too ill to complete the interview and those where the baby died before the interview could take place. At some of the hospitals, parents who were under 18 were not eligible to be interviewed due to hospital prohibition. Children were followed from birth until age fifteen. At the time of birth, medical record information was extracted from the hospital at which the mother gave birth. Interviews were conducted with the mother at birth in person and with father in person or over the telephone if he was not present. Interviews were also conducted with the mother and father when the child was one, three, five, and nine via telephone. Additionally, at ages three, five and nine, a separate primary caregiver survey was administered to the individual with whom the child was living either over the telephone or in the home. The child was interviewed at age nine in the home and at age fifteen by telephone or in the home. At ages nine and fifteen a saliva sample was taken from the child for DNA and DNA methylation measurements [@Reichman:2001wv].

## Prenatal maternal smoking measurement

## Covariates

## DNA methylation measurement

## Sample filtering

## Probe filtering

## Methylation summary measures

We created several methylation summary measures. Global methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set. Methylation clocks were calculated for each sample according to the methods of --INSERT HERE--. To create polymethylation scores, 6,000 (CHECK) CpG sites with regression coefficients reported in @Joubert:2016ue were multiplied by methylation beta values from FFCW and summed. We calculated polymethylation scores using regression coefficients from 4 different regressions reported in @Joubert:2016ue (LIST) and 3 different methods of transforming the methylation beta values (no transformation, mean center, and z-score standardization).

## Statistical analysis

>>>>

---
output:
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
---

## Methods

### Cohort

The Fragile Families and Child Wellbeing Study (FFCW) is a birth cohort of nearly 5,000 children born in 20 U.S. cities between 1998 and 2000 [@Reichman:2001wv]. FFCW sampled births using a three-stage stratified random sample design, described in detail elsewhere, which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv]. FFCW excluded the following births from the study: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, those where the mothers or babies were too ill to complete the interview and those where the baby died before the interview could take place. Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; additional follow up is ongoing. Assessments include medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child. At ages nine and fifteen a saliva sample was taken from the child for DNA and DNA methylation measurements [@Reichman:2001wv].

### Covariates and exposure measurement

Demographic and prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth). Demographic variables included maternal income to poverty ratio and child sex. Prenatal maternal alcohol and other drug use were dichotomized to any vs no prenatal maternal alcohol and drug use respectively. Prenatal maternal smoking was ascertained from the categorical interview question

> "During your pregnancy, how many cigarettes did you smoke? Did you smoke...:
>
> 2 or more packs a day
>
> 1 or more but less than 2 packs per day
>
> 	Less than 1 pack a day
>
> 	None

Because few participants reported smoking a pack or more a day, this question was also dichotomized to any vs no prenatal maternal smoking.

Postnatal second-hand smoke exposure was encoded in two variables. To encapsulate general early childhood smoke exposure, we used a binary any postnatal exposure to maternal or primary caregiver smoking at ages 1 or 5 , constructed from questions about maternal or primary caregiver smoking in the age 1 and 5 interviews. To encapsulate recent postnatal smoke exposure, we used a categorical packs per day maternal or primary care giver smoking in the month prior to age 9 and 15 maternal interview.

Principal components of child genetic ancestry were calculated from genetic data (CHIP?) from child saliva samples. In non-stratified models, the first two components of genetic ancestry from principal component analysis run in the entire analytic subset were used, in ancestry-stratified models the first two principal components from principal component analysis run within each strata were used.

Confounder variables included principal components of child genetic ancestry, prenatal maternal alcohol and other drug use.

Precision variables (variables strongly associated with DNA methylation controlled for to decrease standard errors of coefficient of interest) included child sex, cell proportions estimated from DNA methylation data

### DNA methylation measurement

Salivary samples from the children in FFCW were taken at ages nine and fifteen using the Oragene•DNA sample collection kit (DNA Genotek Inc., Ontario). Saliva DNA was extracted manually using DNA Genotek's purification protocol using prepIT L2P. DNA was bisulfite treated and cleaned using the Zymo Research EZ DNA Methylation kit. Samples were randomized and plated across slides by demographic characteristics. Saliva DNA methylation was measured using the Illumina HumanMethylation 450k BeadArray [@Sandoval2011] and imaged using the Illumina iScan system. All samples were run in a single batch to minimize technical variability.

DNA methylation image data (IDAT) were processed in R statistical software (3.5) using the minfi package [@Aryee2014]. The IDAT pairs (n=1811 ????) were read into R and the minfi *preprocessNoob* function was used to normalize dye bias and apply background correction. Further quality control was applied using the ewastools packages [@Heiss2018]. We dropped samples with \>10% of sites have detection p-value \>0.01 (n=43), sex discordance between DNA methylation predicted sex and recorded sex (n=20), and abnormal sex chromosome intensity (n=3). CpG sites were removed if they had detection p-value \>0.01 in 5% of samples (n=26,830) or were identified as cross-reactive (n=27,782) [@Chen2013]. Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].

Global methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set. Pediatric epigenetic age was calculated for each sample. To create polymethylation scores for prenatal maternal smoke exposure, 6,074 CpG sites with regression coefficients reported in Joubert et al. 2016 [@Joubert:2016ue] were multiplied by mean-centered methylation beta values from FFCW and summed. We calculated polymethylation scores using regression coefficients from 4 different regressions reported in Joubert et al. 2016 [@Joubert:2016ue] (sustained smoking exposure and DNA methylation in newborn cordblood with cell-type control, sustained smoking exposure and DNA methylation in newborn cordblood without cell-type control, sustained smoking exposure and DNA methylation in older children peripheral blood, any smoking exposure and DNA methylation in newborn cordblood ).

Epigenome-wide association analyses were run separately at age 9 and age 15 using the linear models and the R package limma [@Ritchie2015]. We also ran epigenome wide association analyses using DNA methylation measures from both time points using linear mixed effect models with a random intercept for child ID. Gap probes were identified using the function *gaphunter* from the R package minfi [@Aryee2014].
